---
title: "hw04"
output: 
    html_document:
      keep_md: yes
      toc: yes
---

First I'll need to load the tidyverse and knitr packages:

```{r}
library(tidyverse)
library(knitr)
```


## Introducing my tasks for hw04

I have imported several data tables into my local environment, but want to create a new dataset to work with that will contain information from both. Specifically, I am interested in the various levels of off-farm paid work and on-farm capital in various Census Divisions (CD) in British Columbia (BC), particularly those with a high number of blueberry farms.

All of the data frames have values for each CD in BC, but they vary in their number of observations per CD (some have many as they are count data across various categories). All data frames have values for both 2011 and 2016. 

## 1. Re-shaping my data

I am going to perform a variation on Activity #2, using Stats Canada Agricultural Census Data instead of the Gapminder data. Note that I first need to save my newly imported dataset as an Rdata file (using ```save(Blueberry_Farms, file="Blueberry_Farms.RData")```), and then load it using the following chunk:

```{r}
load(file = "Blueberry_Farms.RData")
```


I need to do the same with the (1) Off-farm Work and (2) Total Farm Capital dataframes that I will use later on:

```{r}
load(file = "PaidNonFarmWork.RData")
load(file = "TotalFarmCapital.RData")
```



### 1.1 Using ```gather``` to make a longer dataframe

Currently, one of my dataframes has 2 variables (columns) for each year, and I'd like to re-shape it so that it has a single row per year (and one variable for number of "farms reporting blueberries").


```{r}
knitr::kable(Blueberry_Farms)
```

To re-shape this, I'll use the ```gather``` function. I'll also name this new table "bf", arrange the observations by CD_Region, select for the Census Division name, year, and number of blueberry farms.   

```{r}
bf <- Blueberry_Farms %>%
  gather(key = "year", value = "blueberry_farms", "2011":"2016") %>%
  arrange(CD_Region) %>%
  select(CD_Region, year, blueberry_farms) 
  knitr::kable(bf)
```

### 1.2 Make a plot comparing number of blueberry farms by region

While in the hw04 description activity 2 (re-shaping) involves plotting life expectancy in 2 countries to compare (using multiple years of data), here I only have 2 years of data, but am interested to compare multiple regions. Let's figure out which regions have the highest number of blueberry farms.

```{r}
bf %>% 
  arrange(desc(blueberry_farms)) %>% # add desc argument to display highest values first
  knitr::kable()
```


```{r}
bf %>%
  filter(CD_Region %in% c("Fraser Valley", "Greater Vancouver", "Capital", "Comox Valley")) %>%
  ggplot(aes(x = CD_Region, y = blueberry_farms)) +
  geom_col(aes(color=year), position = "dodge") +
  labs(x="Census Division", y="Number of Farms", title="Number of Blueberry Farms in Four Census Divisions in B.C.") +
  theme_bw(base_size=12) +
  theme(axis.title = element_text(size=14),
        plot.title = element_text(size=16, hjust = 0.5),
        legend.position = "right",
        legend.direction = "vertical",
        legend.title = element_text(size=12))
```

It is interesting to see from this plot that the four Census Divisions with highest number of blueberry farms have shown different trends from 2011 and 2016 -- with the Fraser Valley and the Comox Valley showing an increase, and the Capital and Greater Vancouver regions showing a decrease. 

## 2. Preparing additional dataframe for Paid Work Off-farm

I'd also like to take a look at the extent of paid off-farm work in various different CDs. The dataframe for 'PaidNonFarmWork' also has columns for year, so I will again use ```gather``` so that these two dataframes are in the same format.

```{r}
ofw <- PaidNonFarmWork %>%
  gather(key = "year", value = "number_operators", "2011":"2016") %>%
  arrange(CD_Region) %>%
  select(CD_Region, Avg_Hrs_OffFarmWork, year, number_operators) 
```

Now we have a row for each year of data, and I've defined this new dataframe 'ofw'. 

I'd like to define a single measure for each region to indicate the extent of off-farm work. I think that the proportion of farm operators that work 20+ hours a week off of the farm should be a good measure, so I have some additional re-shaping to do...

### 2.1 Using ```spread```

First, I am going to ```spread``` the data so that I can have the number of operators in each 'bin' as variables instead of observations, as I think it will be easier to work with once in that format. 

```{r}
ofw_wide <- ofw %>%
  spread(Avg_Hrs_OffFarmWork, number_operators) %>% View 
```

Success! Only after re-considering my approach, I will not use the wide format. So for now I will shelve `ofw_wide`, but will leave this in here as a record for how I did it.


### 2.2 Using ```mutate``` and ```cumsum``` to define a new variable

I want to create a new variable for the proportion of farm operators in each region that work 20+ hours off of the farm (and I have data for two years - 2011 and 2016).

First I will make a new column and call it hours, but I'll just keep it empty. Then I will filter for the values I want to use to make my new measure (number of operators in each of the categories for 20 hours and more [3 categories total]).  
going to make a new coumn and number it so that it is ordered

```{r}
ofw_20_40 <- ofw %>% 
  mutate(hours = NA) %>% #creating empty column
  filter(!Avg_Hrs_OffFarmWork %in% c("Total Operators", "No off-farm work", "less than 20 hours")) %>% 
  mutate(year = as.factor(year)) %>% #filtering out the values that I don't want (I only want the bins for 20 hours +)
  mutate(CD_Region = as.factor(CD_Region)) %>%  # changing CD_Region variable to a factor to make sure group_by works on it
  mutate(hours = ifelse(Avg_Hrs_OffFarmWork == "No off-farm work", 0, hours)) %>% # assigning new values for the bins
  mutate(hours = ifelse(Avg_Hrs_OffFarmWork == "less than 20 hours", 20, hours)) %>% 
  mutate(hours = ifelse(Avg_Hrs_OffFarmWork == "20 - 29 hours", 29, hours)) %>% 
  mutate(hours = ifelse(Avg_Hrs_OffFarmWork == "30 - 40 hours", 40, hours)) %>% 
  mutate(hours = ifelse(Avg_Hrs_OffFarmWork == "more than 40 hours", 41, hours)) %>% 
  group_by(CD_Region, year) %>%
  arrange(hours) %>%
  group_by(CD_Region, year) %>%
  mutate(cum_operators = cumsum(number_operators)) %>% #creating another new variable to contain the cumulative sum of the number operators in each region by year (for the 3 categories filtered above)
  arrange(CD_Region, year) #to arrange the ordering of the final table
```

In this new dataframe (ofw_20_40), I can see that the new[est] variable (cum_operators) contains the cumulative sum of operators for the 'bins' of 20-29 hours, 30-40 hours, and 40+ hours of off-farm work per week! I only need the third value (the total of the three), so I will filter that out...and call this dataset ofw_40

```{r}
ofw_40 <- ofw_20_40 %>% 
  filter(Avg_Hrs_OffFarmWork == "more than 40 hours")
```


### 2.3 Joining two dataframes

Now to get the ratio for each region by year, I can use the total number of operators (not found in ofw_20_40 but still in the ofw dataframe I created above). Since this is already a row in the dataset, I will create a new dataframe that contains it, and join it with the one I jsut created with the sum of operators who work more than 20 hours off-farm per week. 

```{r}
ofw_tot <- ofw %>% 
  filter(Avg_Hrs_OffFarmWork == "Total Operators") %>% 
  select(CD_Region, year, Avg_Hrs_OffFarmWork, number_operators)
```

And to join them, I should have two dataframes with the same number and types of rows (2 rows for each region, for 2011 and 2016 respectively), so I can use any join function, but will use ```left_join```:

```{r}
left_join(ofw_40, ofw_tot, by = c("CD_Region", "year"))
```

This looks great! I have some extra/poorly named columns, so I will re-name these to reflect the actual values contained in them:

```{r}
left_join(ofw_40, ofw_tot, by = c("CD_Region", "year")) %>% 
  rename(total_operators = number_operators.y, operators_20plus = cum_operators) %>% 
  select(CD_Region, year, total_operators, operators_20plus )
```

One step closer! Now I can use the ```mutate``` function to create a new variable with the proportion using the values in operators_20plus and total_operators:

```{r}
ofw_prop <- left_join(ofw_40, ofw_tot, by = c("CD_Region", "year")) %>% 
  rename(total_operators = number_operators.y, operators_20plus = cum_operators) %>% 
  select(CD_Region, year, total_operators, operators_20plus ) %>% 
  mutate(prop_20plus = operators_20plus/total_operators)
```



## 3. Joining two *more* dataframes

Now that I have re-shaped two dataframes, I will join them! Because I should have the same observations by region and year, I simply want to combine the columns/variables. As such, I will use the ```full_join``` function.

```{r}
blue_off_farm <- full_join(ofw_prop, bf, by = c("CD_Region", "year")) 
```
### 3.1 Plotting them

Now I'm going to use my newly cleaned, shaped and joined datasets to make a plot:

```{r}
blue_off_farm %>% 
  ggplot(aes(prop_20plus, blueberry_farms, colour=CD_Region)) +
  geom_point()
```





