
---
title: "hw02"
output: 
  html_document: 
    keep_md: yes
---

```{r}
#+ setup, include = FALSE
library(knitr)
opts_chunk$set(fig.path = 'figure/scatterplot-', error = TRUE)
```


Fisrt, I'll call up the gapminder dataset and tidyverse package (meta-package that contains dplyr - what we are actually interested in).

```{r}
library(gapminder)
library(tidyverse)
```

## 1. Exploring the data

#### 1.1 Dataset type/class

One of the first questions to ask when working with a new dataset would be, what kind of form/structure does this dataset have? 

Because I have just done some reading about ```dplyr```, I assume (and know from previous assignments) that this is a dataframe, but if I wanted to confirm I could just call up the dataset to see a nicely formatted snippet of the rows and columns and what it is:

```{r}
gapminder
```

Thanks to the tidyverse meta-package, this comes back as a nicely formatted tibble, and confirms we are working with a dataframe.

I can also use the ```typeof``` command to see the type or storage model of the object. This comes back as a ```list```. Upon investigating further and recalling info from previous classes, I can conclude that the gapminder dataset is both a list and a dataframe (a specific type of list). Some more info can be found here: https://stackoverflow.com/questions/15901224/what-is-difference-between-dataframe-and-list-in-r

```{r}
typeof(gapminder)
```


We can can also use the ```class``` command to see what type of data we are working with:

```{r}
class(gapminder)
```

This command confirms that we are working with a table/dataframe. 

#### 1.2 Dataset size & extent

To find out how many **columns**/variables, we simply use the ```ncol``` command, and similarly for the number of **rows**/observations

```{r}
ncol(gapminder)
nrow(gapminder)
```

Another way to get this information all at once (with one command) is the use the ```summarize``` function, but this requires some counting and deciphering to get the same information :

```{r}
summary(gapminder)
```

However, if you really want quick and specific information about the data, using the discrete commands such as ```ncol``` and ```nrow``` could be helpful. 

#### 1.3 Variable types

To isolate one variable in the data frame, you use the $ sign. We can write code to tell us each variable type individually. The following commands tell us that each variable is an integer, apart from ```lifeExp``` and ```gdpPercap```.

```{r}
typeof(gapminder$country)
typeof(gapminder$continent)
typeof(gapminder$year)
typeof(gapminder$lifeExp)
typeof(gapminder$pop)
typeof(gapminder$gdpPercap)
```

## 2. Investigate individual variables

I'd like to take a closer look at lifeExp (quantitative - double) by country (categorical - integer type).

#### 2.1 Possible values and range

First I'll look at the possible values for each category, starting with the possible entries for country. I can use ```levels``` to call up the full list of countries in the dataset, and ```nlevels``` to see the number of possible countries. I won't type the full list, but there are 142 countries in this dataset!

```{r}
levels(gapminder$country)
nlevels(gapminder$country)
```

OK now let's do that for our quantitative (double) variable - ```lifeExp```. Because we can summarize the range of this variable numerically, we will do this diffently. 

```{r}
summary(gapminder$lifeExp)
```

The summary function gives us the minimum value (23.60), the maximum value (82.60), as well as the median and the mean. 

#### 2.2 Spread and distribution 

I cannot make a histogram of country to see the distribution because it is not numeric, however, I can use the ```table``` command to build a contingency table of the counts for each country entry.

```{r}
table(gapminder$country)
```

This shows me that there are 12 entries for each country, which should each correspond to a year in the series. For fun, I will confirm this assumption...

```{r}
list(gapminder$year)
```

Though not the most elegant way of doing it, this confirms that there are 12 unique rows of data for each country, corresponding to distinct years.

Though a bit silly, we can also see this by making a histogram of ```year``` (12 values all with an equal frequency).

```{r}
hist(gapminder$year)
```

Now let's look at the spread or distribution of ```lifeExp``` - only let's do it visually using a histogram. I've gone ahead and looked up some formatting and added lines of central tendency so we can view these values on the histogram (and a legend to explain what they are!).

```{r}
hist(gapminder$lifeExp,
     col = "peachpuff", # now I am really curious to know what other colour options there are...
     border = "black")
abline(v = mean(gapminder$lifeExp), # adding lines for central tendency
	col = "royalblue",
	lwd = 2)
abline(v = median(gapminder$lifeExp),
	col = "red",
	lwd = 2)
legend(x = "topright", # location of legend within plot area
	c("Mean", "Median"),
	col = c("royalblue", "red"),
lwd = c( 2, 2))
```

This allows us to see quite quickly that the histogram is slightly skewed to the left, as there are many entries/rows of data with values between 70 and 75. Using the ```summary``` function we can call up the precise values of the median (60.71), which is slightly higher than the mean (59.47). 


```{r}
summary(gapminder$lifeExp)
```

## 3. Exploring various plot types

#### 3.1 Scatterplots

```{r}
ggplot(gapminder, aes(x = year, y = lifeExp)) # This just sets up the axes by assigning variables -- links certain scale to geometric object

ggplot(gapminder, aes(x = year, y = lifeExp)) + #adding the geom_point sets up the relationship for the scatterplot
  geom_point()

p <- ggplot(gapminder, aes(x = year, y = lifeExp)) # this initializes the scatterplot
```

Let's also look at a scatterplot by two continuous variables to give us the full 'scatter' effect...

```{r}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) 
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) + 
  geom_point()
```
Let's make that scatterplot a bit more visually appealing...

```{r}
p + geom_point() + scale_x_log10() # transforms the scatterplot to log scale
p <- p + scale_x_log10() # re-defining 'p' to integrate the transformation
p + geom_point(aes(color = continent)) # adding colur based on continent
```

After finding some additional notes re making fancy scatterplots, I want to add 

```{r}
p + geom_point() + scale_x_log10() # transforms the scatterplot to log scale
p <- p + scale_x_log10() # re-defining 'p' to integrate the transformation
p <- p + aes(color = continent) + geom_point() + geom_smooth(lwd = 3, se = FALSE)
p + ggtitle("Life expectancy over time by continent") # adding a title (not sure why left justified)
```

#### 3.2 Bar charts

```{r}
ggplot(gapminder, aes(x = continent)) + geom_bar()

```

#### 3.3 Density plot

```{r}
ggplot(subset(gapminder, continent != "Oceania"), #removing Oceania to simplify because only 2 countries...
              aes(x = lifeExp, color = continent)) + geom_density()
```

#### 3.4 Boxplots (1 categorical, 1 quantitative)

```{r}
ggplot(gapminder, aes(x = continent, y = lifeExp)) + geom_boxplot(aes(colour=continent))
```




## 4. Using `filter()` and `select()`

I recently listened to a podcast about gender equity and political changes post-genocide in Rwanda. As such, I'd like to have  a closer look at gdpPercap and lifeExp in Rwanda chronologically. First I'll use the ```filter``` function to pull up the data from Rwanda and see what we're working with.

#### 4.1 Life Expectancy in Rwanda

```{r}
filter(gapminder, country=="Rwanda")
```

Now, I'll use the filter function and piping to make a scatterplot of life expectancy over time. 

```{r}
gapminder %>% 
  filter(country == "Rwanda") %>%
ggplot(aes(x = year, y = lifeExp)) + 
   geom_point() + # no need for alpha transparency
  geom_smooth(se=FALSE) #removing standard error and allowing for curvy line
```

#### 4.2 Life Expectancy in Africa

It might be interesting to look at life expectancy in other/all countries in Africa during a smaller subset of years...

```{r fig.width=12}
gapminder %>% # altered figure width so we could see the lines better
  filter(continent == "Africa", year >= 1970) %>%
  ggplot(aes(year, lifeExp, group=country, colour=country)) +
  geom_point(alpha=0.3) +
  geom_line()
```

While this is a lot to digest/see, we can clearly see a few outliers in the data at the upper and lower ends

#### 4.3 Life expectancy in Rwanda, Afghanistan, Haiti & Ireland

Then maybe it would be nice to take a look at a few coutries that may show interesting temporal changes in life expectancy. To do this I'd like to call up the data for 4 countries from 4 different continents, and take a closer look at some years more recently...

```{r}
gapminder %>% 
    filter(country %in% c("Rwanda", "Afghanistan", "Haiti", "Ireland"), 
           year >= 1970) %>% 
    select(country, year, lifeExp)
```

Now we can look at these changes for each country over time graphically using the ggplot functions.

```{r}
gapminder %>% 
    filter(country %in% c("Rwanda", "Afghanistan", "Haiti", "Ireland")) %>% 
  ggplot(aes(year, lifeExp, colour=country)) +
  geom_point() +
  geom_line()
```
From this we can see some clear differences between these four countries,and notably the swift and severe decline of life expectancy in Rwanda (corresponding to the 1994 genocide). Note that this minimum is actually in 1992 - there is no data for 1994 (and likely the true minimum is not captured).

## 5. But I want to do more!

Funnily enough, I had not seen this part of the homework before making my selection (or filters, rather...) of countries above.


```{r}
filter(gapminder, country == c("Rwanda", "Afghanistan"))
```

The analyst did succeed in calling up some of the data entries for Rwanda and Afghanistan from the Gapminder dataset. However, these do not seem to be all of the entries for each of these countries. Instead, the analyst has gotten one unique entry for each year, with half corresponding to each of the two countries. I have queried this command (`combine`), but can't quite reason why it returns unique values to each year... look forward to hearing the answer!


