
---
title: "hw02"
output: 
  html_document: 
    keep_md: yes
---
# hw02 RMarkdown File

Fisrt, I'll call up the gapminder dataset and tidyverse package (meta-package that contains dplyr - what we are actually interested in).

```{r}
library(gapminder)
library(tidyverse)
```

## Smell testing the data

One of the first questions to ask when working with a new dataset would be, what kind of form/structure does this dataset have? 

Because I have just done some reading about ```dplyr```, I assume (and know from previous assignments) that this is a dataframe, but if I wanted to confirm I could just call up the dataset to see a nicely formatted snippet of the rows and columns and what it is:

```{r}
gapminder
```

Thanks to the tidyverse meta-package, this comes back as a nicely formatted tibble, and confirms we are working with a dataframe.

I can also use the ```typeof``` command to see the type or storage model of the object. This comes back as a ```list```. Upon investigating further and recalling info from previous classes, I can conclude that the gapminder dataset is both a list and a dataframe (a specific type of list). Some more info can be found here: https://stackoverflow.com/questions/15901224/what-is-difference-between-dataframe-and-list-in-r

```{r}
typeof(gapminder)
```


We can can also use the ```class``` command to see what type of data we are working with:

```{r}
class(gapminder)
```

This command confirms that we are working with a table/dataframe. 

To find out how many **columns**/variables, we simply use the ```ncol``` command, and similarly for the number of **rows**/observations

```{r}
ncol(gapminder)
nrow(gapminder)
```

Another way to get this information all at once (with one command) is the use the ```summarize``` function, but this requires some counting and deciphering to get the same information :

```{r}
summary(gapminder)
```

However, if you really want quick and specific information about the data, using the discrete commands such as ```ncol``` and ```nrow``` could be helpful. 
