
---
title: "hw02"
output: 
  html_document: 
    keep_md: yes
---
# hw02 RMarkdown File

Fisrt, I'll call up the gapminder dataset and tidyverse package (meta-package that contains dplyr - what we are actually interested in).

```{r}
library(gapminder)
library(tidyverse)
```

## Smell testing the data

One of the first questions to ask when working with a new dataset would be, what kind of form/structure does this dataset have? 

Because I have just done some reading about ```dplyr```, I assume (and know from previous assignments) that this is a dataframe, but if I wanted to confirm I could just boldly call up the dataset to see a nicely formatted snippet of the rows and columns and what it is:

```{r}
gapminder
```

Thanks to the tidyverse meta-package, this comes back as a nicely formatted tibble, and confirms we are working with a dataframe.

We can can also use the ```class``` command to see what type of data we are working with:

```{r}
class(gapminder)
```

This command confirms that we are working with a table/dataframe. 

To find out how many variables or columns, we simply use the ```ncol``` command, and similarly for the number of rows/observations

```{r}
ncol(gapminder)
nobvs(gapminder)
```

