---
title: "Gapminder_Data_hw01"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Gapminder Dataset for hw01

## Load packages

Assuming both the gapminder and tidyverse packages are installed, load them using the following commands:

```{r}
library(gapminder)
library(tidyverse)
```

## Explore dataset

Using the tidyverse meta-package, we can view the gapminder data frame as a 'tibble', and have a cleaner view of the dataset. 'names' returns all of the names in the dataframe, 'head' returns the first parts of the data frame, etc.

```{r}
gapminder
names(gapminder)
head(gapminder)
tail(gapminder)
glimpse(gapminder)
```

To obtain a basic statistical overview:

```{r}
summary(gapminder)
```
### Make some plots

Using base R graphics, make some figures!

```{r}
plot(lifeExp ~ year, gapminder)
plot(lifeExp ~ gdpPercap, gapminder)
plot(lifeExp ~ log(gdpPercap), gapminder)
```

## Isolate a variable in the frame

To isolate one variable in the data frame, you use the $ sign. 

### Life Expectancy

```{r}
head(gapminder$lifeExp)
summary(gapminder$lifeExp)
hist(gapminder$lifeExp)
```

### Distribution of observations across continents

```{r}
table(gapminder$continent)
barplot(table(gapminder$continent))
```

## Blindly make figures (for now)

```{r}
p <- ggplot(filter(gapminder, continent != "Oceania"),
            aes(x = gdpPercap, y = lifeExp)) # just initializes
p <- p + scale_x_log10() # log the x axis the right way
p + geom_point() # scatterplot
p + geom_point(aes(color = continent)) # map continent to color
p + geom_point(alpha = (1/3), size = 3) + geom_smooth(lwd = 3, se = FALSE)
p + geom_point(alpha = (1/3), size = 3) + facet_wrap(~ continent) +
  geom_smooth(lwd = 1.5, se = FALSE)
  ```
