---
title: "hw05"
output: 
  html_document: 
    keep_md: yes
---

## 0. Pre-amble

I'm going to use both the gapminder dataset and a dataset from the Canadian Census of Agriculture to go through the exercises for hw05.

First I'll load all of the packages I'm going to need for this assignment in one tidy chunk of code (including <a href="https://github.com/karthik/wesanderson">a new colour palette</a>...)

```{r}
library("wesanderson")
library(gapminder)
library(tidyverse)
library(knitr)
options(knitr.table.format = 'markdown')
library(forcats)
library(readxl) #package already downloaded (bundled in the tidyverse package!) for reading excel files
library(RColorBrewer)
```


## 1. Factor Management

Starting with the gapminder dataset, I'll refresh my memory of what I'm working with here...

```{r}
gapminder
```


As I know this assignment works with factors, I've noted which variables are factors.

I want to drop observations for Oceania, and then drop the unused factor levels (left by Oceania's absense). First let me take a look at the number of factor levels (and associated number of observations for each) there are for the variable ```Continent```:

### 1.1 Dropping Oceania from Gapminder

```{r}
nlevels(gapminder$continent) # total # levels for continent (factor)
fct_count(gapminder$continent) # number of obs per factor using forcats package
```


```{r}
gap_5 <- gapminder %>% 
  filter(!continent == "Oceania")
```

Now I'll use the forcats package to drop unused levels

```{r}
gap_5_dropped <- gap_5 %>% 
  droplevels()
nlevels(gap_5_dropped$continent)
```
 
 Checking to see if this worked...

```{r}
nlevels(gap_5_dropped$continent) # total # levels for continent (factor)
fct_count(gap_5_dropped$continent) # number of obs per factor using forcats package
```

Success!

### 1.2 Using factors in BC Agriculture data

I'll first bring in and take a look at these 2 additional dataset (stored as an excel file in my working directory)

```{r}
exp06 <- read_excel("2006_Expenses.xls")
rec06 <- read_excel("2006_Receipts.xls")
fru06 <- read_excel("2006_Fruits.xls")
View(exp06)
View(rec06)
View(fru06)
```

These are somewhat large files (2087 observations each and between 17 and 52 variables. So I'd like to zoom in on the measures and regions I am most interested in (see <a href="">my README.md file</a> for more details.

#### 1.2.1 Prepping the data

I'll use a combination of select and filter to create new dataframes with the information I care to look at (variables in the readme file and the Census Agricultural Regions (CARs) in BC), and then I'll do some work with factors. 

```{r}
rec06_BC <- rec06 %>% # Dataframe for gross farm receipts
  select(HDPROV, AGOPCAR, CENSUSD, AREANAME, VAL_SALESXFP) %>% #leaving enough information for future analyses
  filter(CENSUSD == 0, AREANAME %in% c("Vancouver Island-Coast", "Lower Mainland-Southwest", "Thompson-Okanagan", "Kootenay", "Cariboo", "North Coast", "Nechako", "Peace River"))

exp06_BC <- exp06 %>%  #Dataframe for total farm expenses
  select(HDPROV, AGOPCAR, CENSUSD, AREANAME, TOTEXP) %>% 
  filter(CENSUSD == 0, AREANAME %in% c("Vancouver Island-Coast", "Lower Mainland-Southwest", "Thompson-Okanagan", "Kootenay", "Cariboo", "North Coast", "Nechako", "Peace River"))

fru06_BC <- fru06 %>% 
  select(HDPROV, AGOPCAR, CENSUSD, AREANAME, BLUEBTA) %>% 
  filter(AREANAME %in% c("Vancouver Island-Coast", "Lower Mainland-Southwest", "Thompson-Okanagan", "Kootenay", "Cariboo", "North Coast", "Nechako", "Peace River"))
```

Great! Much more manageable. Now I'll look at the dataframes to see how the AGOCAR numbers (a categorical variable) and AREANAME are stored.

```{r}
str(exp06_BC$AREANAME)
str(rec06_BC$AGOPCAR)
```

AREANAME is a character, as would be expected, and AGOPCAR is stored as numerical. I'd like to store the AGOPCAR as a factor (since it corresponds to the categorical variable of AREANAME). 

```{r}
rec06_BC$AGOPCAR <- as_factor(as.character(rec06_BC$AGOPCAR)) 
```

Now I'd like to see if I can rename the factor with the CAR names...

```{r}
levels(rec06_BC$AGOPCAR)<-c("Vancouver Island-Coast", "Lower Mainland-Southwest", "Thompson-Okanagan", "Kootenay", "Cariboo", "North Coast", "Nechako", "Peace River")
```

Success (and I can see that the names are accurate because I still have the AREANAME column). 

Now I'd like to do the same with the expenses data...

```{r}
exp06_BC$AGOPCAR <- as_factor(as.character(exp06_BC$AGOPCAR))
levels(exp06_BC$AGOPCAR)<-c("Vancouver Island-Coast", "Lower Mainland-Southwest", "Thompson-Okanagan", "Kootenay", "Cariboo", "North Coast", "Nechako", "Peace River")
```

And now I'll join these two dataframes and drop unnecessary variables..,

```{r}
rec_exp_06_BC <- left_join(exp06_BC, rec06_BC) %>% 
  select(HDPROV, AGOPCAR, TOTEXP, VAL_SALESXFP)
```

Great! Now this is ready for some analysis... 


#### 1.2.2 Re-ordering the factors in a principled way

Check default ordering..
```{r}
rec_exp_06_BC$AGOPCAR %>% 
  levels()
```

I'd like to calculate the percentage of sales that are profit for each region as a summary statistic, then re-order the data based on this. However, I have realized that my variables of TOTEXP and VAL_SALESXFP are currently in the character format! Not going to work for these calculations. So first I'll convert these to numerical.

```{r}
rec_exp_06_BC$TOTEXP <- as.numeric(rec_exp_06_BC$TOTEXP) # Converting to numeric
str(rec_exp_06_BC$TOTEXP) # Checking to be sure it worked
rec_exp_06_BC$VAL_SALESXFP <- as.numeric(rec_exp_06_BC$VAL_SALESXFP)
str(rec_exp_06_BC$VAL_SALESXFP)
```

```{r}
fct_reorder(rec_exp_06_BC$AGOPCAR, rec_exp_06_BC$VAL_SALESXFP, .desc = TRUE) %>% 
  levels()
```

Now I will plot the total gross farm receipts (sales) by the ordered CARs. Only, the colour palette I have been dying to use only has 5 colours (wohn-wohn), so I am going to only plot 5 of the regions.

```{r}
rec_exp_06_BC %>% 
  filter(AGOPCAR %in% c("Lower Mainland-Southwest", "Thompson-Okanagan", "Vancouver Island-Coast", "Peace River", "Cariboo")) %>% 
  ggplot(aes(x = fct_reorder(AGOPCAR, VAL_SALESXFP, .desc = TRUE), y = TOTEXP)) +
  geom_col(aes(fill = AGOPCAR)) +
  labs(x="Census Agricultural Region", y="Total Farm Operating Expenses", title="Total Operating Expenses on Farms by Census Agricultural Region in B.C.") +
  scale_fill_manual(values = wes_palette("Cavalcanti")) + 
  theme(axis.title = element_text(size=14),
        plot.title = element_text(size=16, hjust = 0.5),
        legend.position = "right",
        legend.direction = "vertical",
        legend.title = element_text(size=12))
```

Success!

Now, to make sure I accomplish the tasks from the homework, I'll create a new variable (instead of a summary statistic) that takes into account both measures of sales and expenses ((expenses / sales) * 100).

```{r}
rec_exp_06_BC %>% 
  mutate(perc_exp =(TOTEXP/VAL_SALESXFP)*100) %>% 
  knitr::kable()
```

Now I'll re-order based on this new variable, and plot that...

```{r}
rec_exp_06_BC %>% 
  filter(AGOPCAR %in% c("Lower Mainland-Southwest", "Thompson-Okanagan", "Vancouver Island-Coast", "Peace River", "Cariboo", "Kootenay", "Nechako")) %>% 
  mutate(perc_exp = (TOTEXP/VAL_SALESXFP)*100) %>% 
  ggplot(aes(x = fct_reorder(AGOPCAR, perc_exp, .desc = TRUE), y = perc_exp)) +
  geom_col(aes(fill = AGOPCAR)) +
  labs(x="Census Agricultural Region", y="Operating Expenses as Percent of Farm Receipts", title="Expenses as a Percent of Farm Sales in B.C.") +
  scale_fill_manual(values = wes_palette("BottleRocket")) + 
  theme(axis.title = element_text(size=14),
        plot.title = element_text(size=16, hjust = 0.5),
        legend.position = "right",
        legend.direction = "vertical",
        legend.title = element_text(size=12))
```

I wasn't able to also order the legend.



## 2. Writing to file, Reading

I'm going to try writing the re-ordered dataframe to a csv file.

```{r}
fct_reorder(rec_exp_06_BC$AGOPCAR, rec_exp_06_BC$VAL_SALESXFP, .desc = TRUE) %>% 
  write_csv("order_sales_06.csv")
```
Using the ```fct_reorder``` function won't allow me to do this, as it doesn't re-order the whole dataframe in this case. 


I found this other way to re-order by the dataframe using indexing:

```{r}
rec_exp_06_BC[order(rec_exp_06_BC$VAL_SALESXFP, decreasing = TRUE),] %>% # Keep columns the way they are but order the rows by Sales
  write_csv("rec_exp_06_BC.csv")
```

Now I'll bring it back in...

```{r}
read_csv("rec_exp_06_BC.csv")
```

While the csv file is still in the specified order, the AGOPCAR variable is now a character, once again. So some information has been lost.

## 3. Visualization Design


## 4. Writing figures to file



