---
title: "hw05"
output: 
  html_document: 
    keep_md: yes
---

## 0. Pre-amble

First I'll load all of the packages I'm going to need for this assignment in one tidy chunk of code.

```{r}
library(gapminder)
library(tidyverse)
library(knitr)
options(knitr.table.format = 'markdown')
library(forcats)
library(readxl) #package already downloaded (bundled in the tidyverse package!) for reading excel files
```

I'm going to use the gapminder dataset to go through the exercises for hw05. First I'll refresh my memory of what I'm working with here...

```{r}
gapminder
```


As I know this assignment works with factors, I've noted which variables are factors.


## 1. Factor Management

I want to drop observations for Oceania, and then drop the unused factor levels (left by Oceania's absense). First let me take a look at the number of factor levels (and associated number of observations for each) there are for the variable ```Continent```:

### 1.1 Dropping Oceania from Gapminder

```{r}
nlevels(gapminder$continent) # total # levels for continent (factor)
fct_count(gapminder$continent) # number of obs per factor using forcats package
```


```{r}
gap_5 <- gapminder %>% 
  filter(!continent == "Oceania")
```

Now I'll use the forcats package to drop unused levels

```{r}
gap_5_dropped <- gap_5 %>% 
  droplevels()
nlevels(gap_5_dropped$continent)
```
 
 Checking to see if this worked...

```{r}
nlevels(gap_5_dropped$continent) # total # levels for continent (factor)
fct_count(gap_5_dropped$continent) # number of obs per factor using forcats package
```