---
title: "hw05"
output: 
  html_document: 
    keep_md: yes
---
First I'll load all of the packages I'm going to need for this assignment in one tidy chunk of code.

```{r}
library(tidyverse)
library(knitr)
options(knitr.table.format = 'markdown')
library(forcats)
library(readxl) #package already downloaded (bundled in the tidyverse package!) for reading excel files
```


Next, I'll use the ```readxl``` package to load the agricultural census data that I'll be working with for this assignment.
```{r}
cap81 <- read_excel("1981_Capital.xls")
exp81 <- read_excel("1981_Expenses.xls")
cap86 <- read_excel("1986_Capital.xls")
exp91 <- read_excel("1991_Expenses.xls")
cap96 <- read_excel("1996_Capital.xls")
exp96 <- read_excel("1996_Expenses.xls")
rec96 <- read_excel("1996_Receipts.xls")
exp01 <- read_excel("2001_Expenses_Receipts.xls")
exp06 <- read_excel("2006_Expenses.xls")
rec06 <- read_excel("2006_Receipts.xls")
```

Great! Files are loaded. Only there are thousands more observations than I need. I'd like to filter for the observation in BC (province 59).


Change province variable to a f
```{r}

```



I'll first try doing this with a smaller subset of the datasets loaded above.

```{r}
bc_cap81 <- filter(cap81$PROV==59)
```



