---
title: "hw03"
date: '2017-10-02'
output: 
    html_document:
      keep_md: yes
      toc: yes
---

### By Susanna Klassen

```{r}
suppressPackageStartupMessages(library(tidyverse))  #Note to inquire/remind myself about the functionality of these options
suppressPackageStartupMessages(library(gapminder))
knitr::opts_chunk$set(fig.width=4, fig.height=3)
```


## Outline

I will accomplish the following (+) tasks in this assignment
1. Look at the spread of GDP per capita within the continents.
2. How is life expectancy changing over time on different continents?
3. Find a few countries with interesting stories and explore them.

```{r}
library(gapminder)
library(tidyverse)
```


### 1. GDP per capita by continent

#### 1.1 Table

To look at the spread of GDP per capita by continent I will use the `groupby` function in conjunction with piping.

```{r}
gapminder %>%
  group_by(continent) %>%
summarize(min(gdpPercap), max(gdpPercap), mean(gdpPercap)) 
```

Thinking there might be a more elegant way of getting this info (all measures of the same variable)

```{r}
gapminder %>%
  group_by(continent) %>%
summarize_each(funs(min, max, mean), gdpPercap) # could also add other variables here if I wanted these same stats for multiple variables
```
That also worked!

#### 1.2 Figure

```{r}
gapminder %>%
  group_by(continent) %>%
  ggplot(aes(x = continent, y = gdpPercap)) + geom_boxplot(aes(colour=continent))
```

This boxplot clearly depicts how the spread of GDP per capita differs between continents, and allows us to get a much better sense of the spread of values more quickly than with the table. For example, the range of values for Africa is quite narrow, versus that of Asia or Europe, both of which exhibit different trends in central tendency. 

### 2. Changes in life expectancy over time by continent

#### 2.1 Table

I'll create a table that uses the `mutate` function to display changes in life expectancy over time by continent. I will also select only the variables I want to display (dropping pop and gdpPercap).

```{r}
gapminder %>%
  group_by(continent) %>%
  select(country, continent, year, lifeExp) %>%
  mutate(lifeExp_change = lifeExp - lifeExp[1])
```

This table doesn't really allow me to see much, actually. If I were to create a new variable that is the average across countries (a strange thing to do in this particular case, but hey...)



#### 2.2 Figure

As I'm wanting to plot two continuous variables for this task, I'll want to use a scatter plot. I'll fit a line to the data to better display the change in life expectancy by continent over time. 

```{r}
gapminder %>%
  group_by(continent) %>%
  select(country, continent, year, lifeExp) %>%
  ggplot(aes(year, lifeExp,
            color = continent)) + 
  geom_point() +
  geom_smooth()
```


### 3. Countries with interesting stories

#### 3.1 Exploring the Americas

My first thought is to explore GDP per capita in Venezuela, which has had a somewhat volatile political and economic history (particularly in the last few yeasrs). But before I dive into a specific country, I might as well look at trends in that continent more broadly to try to pick out some potentially interesting stories (which means I'll need to look more broadly at the Americas).

#### 3.1.1 Life expectancy in the Americas


```{r fig.width=12, fig.height=8}
gapminder %>% # altered figure width so we could see the lines better
  filter(continent == "Americas") %>%
  ggplot(aes(year, lifeExp, group=country, colour=country)) +
  geom_point(alpha=0.3) +
  geom_line()
```

Some interesting stories in there.
1. A drastic increase over this time period for some!
2. Canada consistently showing the highest value
3. The United States drops below some other countries later in this time period

#### 3.1.2 GDP per capita in the Americas

```{r fig.width=12, fig.height=8}
gapminder %>% # altered figure width so we could see the lines better
  filter(continent == "Americas") %>%
  ggplot(aes(year, gdpPercap, group=country, colour=country)) +
  geom_point(alpha=0.3) +
  geom_line()
```

Let's drop off Canada and the US to get a closer look at Central & South America (+ Mexico)

```{r fig.width=12, fig.height=8}
gapminder %>% # altered figure width so we could see the lines better
  filter(continent == "Americas", country != "Canada", country != "United States") %>%
  ggplot(aes(year, gdpPercap, group=country, colour=country)) +
  geom_point(alpha=0.3) +
  geom_line() 
```

#### 3.2 Comparing changes in GDP per capita in 5 countries

It was a bit difficult to 'zoom' in on a few of the more interesting patterns in the figure above, but using some trial and error (I look forward to learning a more efficient way of doing this!), I found that Venezuela, Haiti, Puerto Rico, Trinidad and Tobago, and Nicaragua all exhibited interesting trends in GDP per capita over time. 

```{r fig.width=12, fig.height=8}
gapminder %>% # altered figure width so we could see the lines better
  filter(country %in% c("Venezuela", "Haiti", "Puerto Rico", "Trinidad and Tobago", "Nicaragua")) %>%
  ggplot(aes(year, gdpPercap, group=country, colour=country)) +
  geom_point(alpha=0.3) +
  geom_line() 
```


```{r fig.width=12, fig.height=8}
gapminder %>% 
  filter(country %in% c("Venezuela", "Haiti", "Puerto Rico", "Trinidad and Tobago", "Nicaragua")) %>% 
  ggplot(aes(gdpPercap, lifeExp, group=country, colour=country)) +
  facet_wrap(~ country) +
  geom_point(alpha=0.2) +
  geom_line()
```
This was an imperfect attempt to display trends in GDP per capita and life expectancy (2 continuous variables) in these 5 countries using faceting. It shows something, but maybe isn't a fully successful attempt. Part of the issue here is the common x scale. Let's try reversing the x and y variables...

```{r fig.width=12, fig.height=8}
gapminder %>% 
  filter(country %in% c("Venezuela", "Haiti", "Puerto Rico", "Trinidad and Tobago", "Nicaragua")) %>% 
  ggplot(aes(lifeExp, gdpPercap, group=country, colour=country)) +
  facet_wrap(~ country) +
  geom_point(alpha=0.2) +
  geom_line()
```

Still kind of a convoulted way to look at these variables...

All in one plot:

```{r fig.width=12, fig.height=8}
gapminder %>% 
  filter(country %in% c("Venezuela", "Haiti", "Puerto Rico", "Trinidad and Tobago", "Nicaragua")) %>% 
  ggplot(aes(lifeExp, gdpPercap, colour=country)) +
    geom_point() +
    geom_smooth()
```

#### 3.3 Contrasting Nicaragua and Puerto Rico

These two nations are interesting to contrast, as they display very different trends in life expectancy, but are geographically close to one another (which doesn't necessarily mean anything, but alas, it is interesting).

Let's look more closely at the change in GDP per capita in Puerto Rico...

```{r}
gapminder %>%
  filter(country == "Puerto Rico") %>% 
  select(country, continent, year, pop, gdpPercap)
```

And comparing changes in GDP per capita over time for both countries...

```{r}
gapminder %>% 
  filter(country %in% c("Puerto Rico","Nicaragua")) %>% 
ggplot(aes(year, gdpPercap, color = country)) + 
  geom_point() +
  geom_smooth()
```

Let's see how they compare when we factor in changes in population...

```{r}
gapminder %>% 
  filter(country %in% c("Puerto Rico","Nicaragua")) %>% 
ggplot(aes(pop, gdpPercap, color = country)) + 
  geom_point() +
  geom_smooth()
```

Wow! That is super interesting... it looks like they have experienced very different trends in population growth... let's visualize that to compare them.

```{r}
gapminder %>% 
  filter(country %in% c("Puerto Rico","Nicaragua")) %>% 
ggplot(aes(year, pop, color = country)) + 
  geom_point() +
  geom_smooth()
```

Puerto Rico has seen more gradual population growth, starting with a higher population, but Nicaragua's population surpassed it around 1987. This shows that growth in GDP in Puerto Rico must have outpaced population growth (as the GDP per capita value becomes quite high later on in this time period.)


