---
title: "hw03"
date: '2017-10-02'
output: 
    html_document:
      keep_md: yes
      toc: yes
---

### By Susanna Klassen

```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
knitr::opts_chunk$set(fig.width=4, fig.height=3)
```


## Outline

I will accomplish the following (+) tasks in this assignment
1. Look at the spread of GDP per capita within the continents.
2. How is life expectancy changing over time on different continents?
3. Find a few countries with interesting stories and explore them.

```{r}
library(gapminder)
library(tidyverse)
```


### 1. GDP per capita by continent

#### 1.1 Table

To look at the spread of GDP per capita by continent I will use the `groupby` function in conjunction with piping.

```{r}
gapminder %>%
  group_by(continent) %>%
summarize(min_gdpPercap=min(gdpPercap), max_gdpPercap=max(gdpPercap), mean_gdpPercap=mean(gdpPercap)) 
```



#### 1.2 Figure

```{r}
gapminder %>%
  group_by(continent) %>%
  ggplot(aes(x = continent, y = gdpPercap)) + geom_boxplot(aes(colour=continent))
```

This boxplot clearly depicts how the spread of GDP per capita differs between continents, and allows us to get a much better sense of the spread of values more quickly than with the table. For example, the range of values for Africa is quite narrow, versus that of Asia or Europe, both of which exhibit different trends in central tendency. 

### 2. Changes in life expectancy over time by continent

#### 2.1 Table

I'll create a table that uses the `mutate` function to display changes in life expectancy over time by continent. I will also select only the variables I want to display (dropping pop and gdpPercap).

```{r}
gapminder %>%
  group_by(continent) %>%
  select(country, continent, year, lifeExp) %>%
  mutate(lifeExp_change = lifeExp - lifeExp[1])
```

This table doesn't really allow me to see much, actually. If I were to create a new variable that is the average across countries (a strange thing to do in this particular case, but hey...)

```{r}
gapminder %>%
  arrange(year) %>%
  group_by(continent) %>%
  mutate(lifeExp_change = lifeExp[year==2007] - lifeExp[year==1997])
```

#### 2.2 Figure

As I'm wanting to plot two continuous variables for this task, I'll want to use a scatter plot. I'll fit a line to the data to better display the change in life expectancy by continent over time. 

```{r}
gapminder %>%
  group_by(continent) %>%
  select(country, continent, year, lifeExp) %>%
  ggplot(aes(year, lifeExp, color = continent)) + 
  geom_point() +
  geom_smooth()
```
This scatterplot aggregated at the scale of a continent is a fairly simple figure both to produce and interpret.

### 3. Countries with interesting stories

#### 3.1 Exploring the Americas

My first thought is to explore GDP per capita in Venezuela, which has had a somewhat volatile political and economic history (particularly in the last few years). But before I dive into a specific country, I might as well look at trends in that continent more broadly to try to pick out some potentially interesting stories.

#### 3.1.1 Life expectancy in the Americas

Let's look at the average value of life expectancy over the last 20 years of the dataset by country (in the Americas). I am going to assign an object (t3.1) to this table, then plot it.

```{r}
t3.1 <- gapminder %>%
  filter(continent == "Americas", year >= 1987) %>%
  group_by(country)%>%
  summarize(mean_lifeExp=mean(lifeExp))
```

Test that to see if it worked...
```{r}
t3.1
```

Now let's plot that:

```{r}
t3.1 + ggplot() +
  geom_point()
```


```{r fig.width=12, fig.height=8}
gapminder %>% # altered figure width (12) and height (8) so we could see the lines better
  filter(continent == "Americas") %>%
  ggplot(aes(year, lifeExp, group=country, colour=country)) +
  geom_point(alpha=0.3) +
  geom_line()
```

While it is difficult to see which colorus are which (the legend is only marginally helpful), there are some interesting stories in there.
1. A sizeable increase over this time period for most countries!
2. Canada consistently showing the highest value
3. The United States drops below some other countries later in this time period

#### 3.1.2 GDP per capita in the Americas

```{r fig.width=12, fig.height=8}
gapminder %>% 
  filter(continent == "Americas") %>%
  ggplot(aes(year, gdpPercap, group=country, colour=country)) +
  geom_point(alpha=0.3) +
  geom_line()
```

This spaghetti plot is similar to the last - it's difficult to tease out which country is which from the legend, but I can clearly see some outliers in this plot (Canada & US). Let's drop off Canada and the US to get a closer look at Central & South America (+ Mexico)

```{r fig.width=12, fig.height=8}
gapminder %>% # altered figure width so we could see the lines better
  filter(continent == "Americas", country != "Canada", country != "United States") %>%
  ggplot(aes(year, gdpPercap, group=country, colour=country)) +
  geom_point(alpha=0.3) +
  geom_line() 
```

Another option would be to use colours to highlight a few countries of interest (e.g. making all lines black but Venezuela). I'm not sure exactly how to do this at this point and am out of time, but am flagging this as an idea for future exploration. 

#### 3.2 Comparing changes in GDP per capita in 5 countries

It was a bit difficult to 'zoom' in on a few of the more interesting patterns in the figure above, but using some trial and error (I look forward to learning a more efficient way of doing this!), I found that Venezuela, Haiti, Puerto Rico, Trinidad and Tobago, and Nicaragua all exhibited interesting trends in GDP per capita over time. 

```{r fig.width=12, fig.height=8}
gapminder %>% # altered figure width so we could see the lines better
  filter(country %in% c("Venezuela", "Haiti", "Puerto Rico", "Trinidad and Tobago", "Nicaragua")) %>%
  ggplot(aes(year, gdpPercap, group=country, colour=country)) +
  geom_point(alpha=0.3) +
  geom_line() 
```


```{r fig.width=12, fig.height=8}
gapminder %>% 
  filter(country %in% c("Venezuela", "Haiti", "Puerto Rico", "Trinidad and Tobago", "Nicaragua")) %>% 
  ggplot(aes(gdpPercap, lifeExp, group=country, colour=country)) +
  facet_wrap(~ country) +
  geom_point(alpha=0.2) +
  geom_path() #using geom_point instead of geom_line here -- connects in order that they appear
```
This was an imperfect attempt to display trends in GDP per capita and life expectancy (2 continuous variables) in these 5 countries using faceting. It shows something, but maybe isn't a fully successful attempt. Let's try reversing the x and y variables...

Add arrange by year

```{r fig.width=12, fig.height=8}
gapminder %>% 
  filter(country %in% c("Venezuela", "Haiti", "Puerto Rico", "Trinidad and Tobago", "Nicaragua")) %>% 
  ggplot(aes(lifeExp, gdpPercap, group=country, colour=country)) +
  facet_wrap(~ country) +
  geom_point(alpha=0.2) +
  geom_path()
```

Still kind of a convoulted way to look at these variables...

All in one plot:

```{r fig.width=12, fig.height=8}
gapminder %>% 
  filter(country %in% c("Venezuela", "Haiti", "Puerto Rico", "Trinidad and Tobago", "Nicaragua")) %>% 
  ggplot(aes(lifeExp, gdpPercap, colour=country)) +
  geom_point() +
  geom_smooth(aes(fill=country)) 
```

#### 3.3 Contrasting Nicaragua and Puerto Rico

These two nations are interesting to contrast, as they display very different trends in life expectancy, but are geographically close to one another (which doesn't necessarily mean anything, but alas, it is interesting).

Let's look more closely at the change in GDP per capita in Puerto Rico...

```{r}
gapminder %>%
  filter(country == "Puerto Rico") %>% 
  select(country, continent, year, pop, gdpPercap)
```

And comparing changes in GDP per capita over time for both countries...

```{r}
gapminder %>% 
  filter(country %in% c("Puerto Rico","Nicaragua")) %>% 
  ggplot(aes(year, gdpPercap, color = country)) + 
  geom_point() +
  geom_smooth()
```

Let's see how they compare when we factor in changes in population...

```{r}
gapminder %>% 
  filter(country %in% c("Puerto Rico","Nicaragua")) %>% 
  ggplot(aes(pop, gdpPercap, color = country)) + 
  geom_point() +
  geom_smooth()
```

Wow! That is super interesting... it looks like they have experienced very different trends in population growth... let's visualize that to compare them.

```{r}
gapminder %>% 
  filter(country %in% c("Puerto Rico","Nicaragua")) %>% 
ggplot(aes(year, pop, color = country)) + 
  geom_point() +
  geom_smooth()
```

Puerto Rico has seen more gradual population growth, starting with a higher population, but Nicaragua's population surpassed it around 1987. This shows that growth in GDP in Puerto Rico must have outpaced population growth (as the GDP per capita value becomes quite high later on in this time period.)


